---
title: "Catch-at-age"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: FALSE
    code_folding: show
    number_sections: TRUE
---
   
   
# SET-UP
```{r  markdown, echo=F}
library(knitr)
opts_chunk$set(echo = T, collapse = T, fig.align = "center", fig.width = 9, fig.height = 6)
options(width = 140)
```    

## settings
```{r settings, message = F}
source('0.0_settings.R')
dir.create(paste0('../../csv/',tail(my.year,1),'/'),showWarnings = FALSE)
```

## load all data
```{r data, message = F}
load(paste0(dir.rdat, "catch.Rdata"))                                # from 1.0 read
load(paste0(dir.rdat, "bio.Rdata"))                                  # from 1.0 read
load(paste0(dir.rdat, "lf.Rdata"))                                   # from 1.0 read
group.region <- read.table(paste0(dir.dat,'caa_group_region.txt'),header=T)
group.gear <- read.table(paste0(dir.dat,'caa_group_gear.txt'),header=T)
```

# LENGTH-WEIGHT RELATIONSHIPS 

Sample weights are often missing and apparently are not precise. Therefore, the weight of a sample is estimated based on the length of the individuals and their number sampled. This section fits length-weight relationships for all years and periods based on the bio database.

```{r lw, message = F}
## prepping
bio$period <- ceiling(bio$month/3)
bio$weight <- bio$weight *1000          # g instead of kg (idem to carbio before)
lw <- bio[!is.na(bio$length) & !is.na(bio$weight),c('year','period','length','weight')]
lw$logw <- log(lw$weight)           
lw$logl <- log(lw$length)          

## remove extremes
# ggplot(lw,aes(x=logl,y=logw))+geom_point()
lw <- lw[lw$logw>0 & lw$logl>3,]                         # minimal weight(1 g) and length (126mm)

## relationships by year and period (quarter) (!robust regression instead of removing subjectively some points each year)
lw.modR <- paste0(dir.rdat, "lw.mod.Rdata")

if(!file.exists(lw.modR)){
    lw.mod <- with(lw,fit.lw(logl,logw,year,period,tresh=50,period.unit='quarter')) # robust regression
    save(lw.mod, file =  lw.modR)
}else{ 
    load(lw.modR)
}

# Creation time
df <- file.info(list.files(dir.rdat, full.names = T,pattern="lw"))
df[which.max(df$mtime),'mtime']
```

## CHECK {.tabset}
### Prediction
```{r lw_fit, message=FALSE}
expand.grid.df <- function(...) type.convert(Reduce(function(...) merge(..., by=NULL), list(...)))
check <- expand.grid.df(lw.mod,data.frame(length=0:ceiling(max(lf$length))))
check$weight <- with(check,exp(intercept+slope*log(length))*cf)
ggplot(check,aes(x=length,y=weight,col=year,group=year))+
    geom_line(size=0.5)+
    facet_wrap(~period)+
    scale_y_continuous(limits=c(0,max(lw$weight)))+
    scale_color_viridis_c()+
    labs(title='Predicted length-weight relationships by trimester',x='Length (mm)',y='Weight (g)')
```


# LENGTH-FREQUENCY DATA

Note:
Q: There are bio samples with 100+ length measurements (appear random samples) that have NO corresponding lf. Do I need to append these to lf?
Examples: sample ids 20801, 20803, 208010, 208014, 60658,61456, 61556, 61840 (see plotted distributions).
A: No reason to do so. Such samples appear to be be from herring by-catch, or it cannot be excluded that this is the case. Number of fish measured is also lower than average and protocol standard (<150 fish).

## are there LFs that are from by-catch???
## remove lfs with less than x fish. e.g. in 1984 some only 10 or 22.
```{r lf, message = F, fig.height = 3,fig.width = 4}
lf.caaR <- paste0(dir.rdat, "lf.caa.Rdata")
if(!file.exists(lf.caaR)){
    ## subset
    lf.caa <- lf[!is.na(lf$length),c('year','month','nafo','gear','length','n','source','sample.id')]
    
    ## exclude length-frequency samples from acoustics (96), research (92, 93)(not present)
    table(lf.caa$source)
    lf.caa$source <- NULL
   
    ## group regions and gears (number of NA in table should be 0)
    # PERIOD: quarter
    lf.caa$period <- ceiling(lf.caa$month/3)
    lf.caa$month <- NULL
    table(lf.caa$period,useNA = 'always')
    
    # REGION: large scale regions 
    lf.caa$region <- group.region[match(lf.caa$nafo,group.region$nafo),'region']
    lf.caa$nafo <- NULL
    table(lf.caa$region,useNA = 'always')
    
    # GEAR: similar selectivity gears
    lf.caa$gear <- group.gear[match(lf.caa$gear,group.gear$gear.cat),'gear.group']
    table(lf.caa$gear,useNA = 'always')
    
    ## Append length-frequencies from bio if there are some that were send directly to IML and not added to lf.
    # No reason to do so.

    # remove lf samples with less than 80 fish measured (how are these lfs??)
    n.check <- tapply(lf$n, lf$sample.id, sum)
    paste0(round(length(n.check[n.check<70])/length(n.check)*100,2),"% removed")
    lf.caa <- lf.caa[!lf.caa$sample.id %in% as.numeric(names(n.check[n.check<70])),]
    
    ## convert lengths to sample weights (kg)
    lf.caa <- merge(lf.caa,lw.mod,all.x=T)  
    lf.caa$weight.unit <- with(lf.caa,exp(intercept+slope*log(length))*cf/1000)
    table(lf.caa$regression,useNA = 'always') # assure no NAs
    
    ## save
    save(lf.caa, file =  lf.caaR)
}else{ 
    load(lf.caaR)
}

# Creation time
df <- file.info(list.files(dir.rdat, full.names = T,pattern="lf"))
df[which.max(df$mtime),'mtime']
```

## CHECK {.tabset}
### model samples
```{r lw_table}
check <- unique(lf.caa[,c('year','period','option')])
kable(with(check,table(option,period)))
```

### predictions
```{r lw_predictions, fig.height = 3,fig.width = 10}
grid.arrange(
  ggplot(lf.caa,aes(x=weight.unit))+geom_histogram(binwidth = 0.1)+labs(title='Predicted unit weights',x='Weight (kg)'),
  ggplot(lf.caa,aes(x=weight.unit*n))+geom_histogram(binwidth = 1)+labs(title='Predicted sample weights',x='Weight (kg)'),
  ncol=2
)
```

## TABLES {.tabset}
### overview
```{r lf_tab_overview, message = F}
df <- lf.caa
df$id <- with(df,paste(sample.id,year,period,region,gear),collapse='.')
lf.overall <- ddply(df,c('year'),summarise,
                    N=length(unique(id)),
                    n.fish=sum(n),
                    mean.fish=round(n.fish/N,0),
                    min.fish=min(tapply(n,id,sum)), # small samples: to check!!
                    max.fish=max(tapply(n,id,sum)))
write.csv(lf.overall,paste0('../../csv/',tail(my.year,1),'/lf.csv'),row.names = FALSE)
kable(lf.overall, caption = "Sample info (samples >70fish)")
```

### period
```{r lf_tab_period, message = F}
dfp <- ddply(df,c('year','period'),summarise,N=length(unique(id)),n=sum(n))
dfp <- dcast(dfp,year~period,value.var='N',fill = 0)
dfp$total <- rowSums(dfp[,-1])
kable(dfp)
```

### region
```{r lf_tab_region, message = F}
dfr <- ddply(df,c('year','region'),summarise,N=length(unique(id)),n=sum(n))
dfr <- dcast(dfr,year~region,value.var='N',fill = 0)
dfr$total <- rowSums(dfr[,-1])
kable(dfr)
```
### gear
```{r lf_tab_gear, message = F}
dfg <- ddply(df,c('year','gear'),summarise,N=length(unique(id)),n=sum(n))
dfg <- dcast(dfg,year~gear,value.var='N',fill = 0)
dfg$total <- rowSums(dfg[,-1])
kable(dfg)
```

## PLOTS {.tabset}

### period
```{r lf_samples_period, message = F}
dfp <- ddply(df,c('year','period'),summarise,N=length(unique(id)),n=sum(n))
grid.arrange(ggplot(dfp,aes(x=year,y=N,fill=as.factor(period)))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(fill='Period')+labs(x=''),
             ggplot(dfp,aes(x=year,y=n,fill=as.factor(period)))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(fill='Period')+labs(x='Year'),ncol=1)
```

### region
```{r lf_samples_region, message = F}
dfr <- ddply(df,c('year','region'),summarise,N=length(unique(id)),n=sum(n))
grid.arrange(ggplot(dfr,aes(x=year,y=N,fill=region))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x=''),
             ggplot(dfr,aes(x=year,y=n,fill=region))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='Year'),ncol=1)

```

### gear
```{r lf_samples_gear, message = F}
dfg <- ddply(df,c('year','gear'),summarise,N=length(unique(id)),n=sum(n))
grid.arrange(ggplot(dfg,aes(x=year,y=N,fill=gear))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x=''),
             ggplot(dfg,aes(x=year,y=n,fill=gear))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='Year'),ncol=1)

```

# BIO DATA
```{r al global, message = F, fig.height = 6,fig.width = 14}
## subset
bio.caa <- bio[!is.na(bio$agef) & !is.na(bio$length),
               c('year','period','nafo','gear','length','agef','sample.id')]
names(bio.caa)[6] <- 'age'

## remove absurd ones
# ages too high
table(bio.caa$age)
bio.caa <- bio.caa[bio.caa$age<18,]    # 4 out of 6 fish have wrong length, very unsure about remaining two

# length not matching
bio.caa <- ddply(bio.caa,c('age'),transform,outlier=outlier(length,coef=3)) # rule of thumb extremes
bio.caa[bio.caa$age==0 & bio.caa$length>300,'outlier'] <- TRUE              # odd second mod
table(bio.caa$outlier,bio.caa$age)

# check
grid.arrange(
  ggplot(bio.caa,aes(x=length,fill=outlier))+
    geom_histogram()+
    facet_wrap(~age,scale='free'),
  ggplot(bio.caa,aes(x=age,y=length,group=age))+
    geom_boxplot()+
    geom_point(data=bio.caa[bio.caa$outlier==T,],aes(col=outlier)),
  ncol=2,widths=c(2,1)
)
bio.caa <- bio.caa[bio.caa$outlier==FALSE,]
bio.caa$outlier <- NULL

## group regions and gears (number of NA in table should be 0)
# PERIOD: quarter
table(bio.caa$period,useNA = 'always')

# REGION: large scale regions 
bio.caa$region <- group.region[match(bio.caa$nafo,group.region$nafo),'region']
bio.caa$nafo <- NULL
table(bio.caa$region,useNA = 'always')

# GEAR: similar selectivity gears
bio.caa$gear <- group.gear[match(bio.caa$gear,group.gear$gear.cat),'gear.group']
table(bio.caa$gear,useNA = 'always')

## make equal length classes
bio.caa$length <- roundFish(bio.caa$length,5) 
bio.caa <- bio.caa[bio.caa$length>0,]
range(bio.caa$length)

save(bio.caa, file =  paste0(dir.rdat, "bio.caa.Rdata"))
```

## TABLES {.tabset}

### overview
```{r al_tab_global, message = F}
df <- bio.caa
df$id <- with(df,paste(sample.id,year,period,region,gear),collapse='.')
al.overall <- ddply(df,c('year'),summarise,
                    N=length(unique(id)),
                    n=length(id),
                    mean.fish=round(mean(table(id)),0),
                    min.fish=min(tapply(age,id,length)), # small samples: to check!!
                    max.fish=max(tapply(age,id,length)))
write.csv(al.overall,paste0('../../csv/',tail(my.year,1),'/al.csv'),row.names = FALSE)
kable(al.overall)
```

### period
```{r al_tab_period, message = F}
dfp <- ddply(df,c('year','period'),summarise,N=length(unique(id)),n=length(age))
dfp <- dcast(dfp,year~period,value.var='N',fill = 0)
dfp$total <- rowSums(dfp[,-1])
kable(dfp)
```

### region
```{r al_tab_region, message = F}
dfr <- ddply(df,c('year','region'),summarise,N=length(unique(id)),n=length(age))
dfr <- dcast(dfr,year~region,value.var='N',fill = 0)
dfr$total <- rowSums(dfr[,-1])
kable(dfr)
```

### gear
```{r al_tab_gear, message = F}
dfg <- ddply(df,c('year','gear'),summarise,N=length(unique(id)),n=length(age))
dfg <- dcast(dfg,year~gear,value.var='N',fill = 0)
dfg$total <- rowSums(dfg[,-1])
kable(dfg)
```


## PLOTS {.tabset}
```{r al_samples, message = F}
df <- bio.caa
df$id <- with(df,paste(sample.id,year,period,region,gear),collapse='.')
```

### period
```{r al_samples_period, message = F}
dfp <- ddply(df,c('year','period'),summarise,N=length(unique(id)),n=length(id))
grid.arrange(ggplot(dfp,aes(x=year,y=N,fill=as.factor(period)))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(fill='Period')+labs(x='')+scale_y_continuous(expand=c(0,0)),
             ggplot(dfp,aes(x=year,y=n,fill=as.factor(period)))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(fill='Period')+labs(x='Year')+scale_y_continuous(expand=c(0,0)),ncol=1)
```

### region
```{r al_samples_region, message = F}
dfr <- ddply(df,c('year','region'),summarise,N=length(unique(id)),n=length(id))
grid.arrange(ggplot(dfr,aes(x=year,y=N,fill=region))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='')+scale_y_continuous(expand=c(0,0)),
             ggplot(dfr,aes(x=year,y=n,fill=region))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='Year')+scale_y_continuous(expand=c(0,0)),ncol=1)
```

### gear
```{r al_samples_gear, message = F}
dfg <- ddply(df,c('year','gear'),summarise,N=length(unique(id)),n=length(id))
grid.arrange(ggplot(dfg,aes(x=year,y=N,fill=gear))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='')+scale_y_continuous(expand=c(0,0)),
             ggplot(dfg,aes(x=year,y=n,fill=gear))+geom_bar(stat='identity')+scale_fill_viridis_d()+labs(x='Year')+scale_y_continuous(expand=c(0,0)),ncol=1)
```

# CATCH DATA
```{r catch split, message = F, fig.height = 8,fig.width = 6}
catch.caa <- catch[,c('year','month','nafo','gear.cat','source','catch')]

## identify period, region and gear (NA allowed)
# 1) PERIOD: quarters
catch.caa$period <- ceiling(catch.caa$month/3)
table(catch.caa$month,useNA = 'always')
catch.caa$month <- NA

# 2) REGION: large scale regions 
catch.caa$region <- group.region[match(catch.caa$nafo,group.region$nafo),'region']
table(catch.caa$gear,useNA = 'always')
catch.caa$nafo <-NULL

# 3) GEAR: similar selectivity gears
catch.caa$gear <- group.gear[match(catch.caa$gear.cat,group.gear$gear.cat),'gear.group']
table(catch.caa$gear,useNA = 'always')

## total catch by level
catch.caa <- ddply(catch.caa,c('year','region','period','gear'),summarise,catch=sum(catch,na.rm = T))
catch.caa <- catch.caa[catch.caa$catch>0,]

save(catch.caa, file =  paste0(dir.rdat, "catch.caa.Rdata"))
```

## PLOTS {.tabset}

### region
```{r catch_region, message = F}
plotCatch(catch.caa,x='Year',y='Landings (t)',fill='region')+scale_fill_viridis_d(na.value = "grey50")
```

### gear
```{r catch_gear, message = F}
plotCatch(catch.caa,x='Year',y='Landings (t)',fill='gear')+scale_fill_viridis_d(na.value = "grey50")
```

### period
```{r catch_period, message = F}
plotCatch(catch.caa,x='Year',y='Landings (t)',fill='period')+scale_fill_viridis_d(na.value = "grey50")
```


# CATCH-AT-AGE
```{r caa, message = F}
catch.diffusedR <- paste0(dir.rdat, "catch.diffused.Rdata")
caaR <- paste0(dir.rdat, "caa.Rdata")
  
if(!all(file.exists(c(catch.diffusedR,caaR)))){
    # step 1) get samples
    lf.caa[,which(names(lf.caa) %in% names(lw.mod)[-c(1:2)])] <- NULL   # remove model details
    
    years <- my.year[my.year>=min(lf$year)]                   # no lf prior to this year
    catch.caa <- catch.caa[catch.caa$year %in% years,]        # years for which lf
    bio.caa <- bio.caa[bio.caa$year %in% years,]              # years for which lf
    
    start_time <- Sys.time()
    catch.diffused <- get.samples(catch=catch.caa,
                       lf=lf.caa, 
                       al=bio.caa,
                       min.lf.samples =2,          # min samples for length-frequency
                       min.al.samples =2,          # min samples for age-length key
                       min.al.fish = 12,      # minimum number of fish for age-length key
                       period.unit='quarter', # quarterly grouping instead of monthly
                       prob.al=0.75,          # see function help
                      parallel=TRUE)        
    end_time <- Sys.time()
    end_time - start_time
    
    catch.diffused$age.0 <- NULL                # remove age 0 (total landings somewhat smaller)
    save(catch.diffused, file =  catch.diffusedR)
    
    # step 2) get actual caa
    caa <- get.caa(x=catch.diffused,plus=10)
    save(caa, file =  caaR)     
}else{ 
    load(catch.diffusedR)
    load(caaR)
}

caa <- round(caa)

write.csv(caa,paste0('../../csv/',tail(my.year,1),'/caa.csv'))

```



## CHECK {.tabset}
### Counts
```{r caa_check_samples, fig.height = 6,fig.width = 8}
plotAttribution(catch.diffused)
```

### %catch per option
```{r caa_check_samples_catch, fig.height = 6,fig.width = 8}
grid.arrange(
    plotQuality(x=catch.diffused,"option.lengthfreq"),
    plotQuality(x=catch.diffused,"option.agelength")
)
```

### %catch per option (lf, annual)
```{r caa_check_samples_catch_annual_lf, fig.height = 8,fig.width = 8}
plotQuality(x=catch.diffused,"option.lengthfreq",annual=TRUE)+geom_hline(yintercept=0.5,col='lightgrey')
```

### %catch per option (alk, annual)
```{r caa_check_samples_catch_annual_alk, fig.height = 8,fig.width = 8}
plotQuality(x=catch.diffused,"option.agelength",annual=TRUE)+geom_hline(yintercept=0.5,col='lightgrey')
```

### total
```{r caa_check_total, fig.height = 6,fig.width = 8}
plotTotal(catch.diffused,caa)
```

## PLOTS {.tabset}
### CAAN
```{r caan_plot, message = F, fig.height = 3,fig.width = 10}
plotCaan(caa)
```

### CAAW
```{r caaw_plot, message = F, fig.height = 3,fig.width = 10}
plotCaaw(caa)
```

### WAA
```{r waa_plot, message = F, fig.height = 4,fig.width = 10}
plotWaa(caa)
```

### WAA.ci
```{r waa_plot_ci, message = F, fig.height = 6,fig.width = 10}
plotWaa(caa,ci=TRUE)
```

### LAA
```{r laa_plot, message = F, fig.height = 4,fig.width = 10}
plotLaa(caa)
```

### LAA.ci
```{r laa_plot_ci, message = F, fig.height = 6,fig.width = 10}
plotLaa(caa,ci=TRUE)
```